---
title: NC Child Dashboard
format: dashboard
execute:
    echo: false
out-width: 100%
fig-width: 7
fig-height: 3
# this will render plots as "vector", meaning they will be sharper
fig-format: svg
fig-dpi: 300
---

```{r}
#| include: false
# ^ this means that any output from this code "chunk" will not be included in 
# the dashboard

# tidyverse is an R package for data manipulation and plotting
library(tidyverse)

# gt is a package for making tables
library(gt)

# tigris is a package for downloading US GIS data
library(tigris)

# ggspatial is a package for making maps
library(ggspatial)

# this will make future "renders" faster by caching map data
options(tigris_use_cache = T)
```

```{r}
#| include: false
# We will make a map of counties in several places, so download GIS data on NC counties once
counties = tigris::counties(state="nc")
```

<!-- Level 1 headings (lines starting with one # sign) become tabs in the navigation bar -->
# Overview {orientation=columns}

<!-- Level 2 headings become columns. The title is not used. -->
## Introduction {width=60%}

This is a reproduction of the [NC Child Data Center](https://ncchild.org/data-center/) in a [Quarto dashboard](https://quarto.org/docs/dashboards/).

## Statewide statistics {width=40%}

### Live births

```{r}
#| content: valuebox
# ^ valuebox will display this as a number in a box

# these data are from https://datacenter.aecf.org/data/tables/2294-live-births#detailed/2/any/false/1096,2545,1095,2048,574,1729,37,871,870,573/any/4792
births = read_csv("data/live_births.csv")

most_recent_year = births |>
    # filter to statewide
    filter(Location == "North Carolina") |>
    # get the row with the largest (most recent) year
    slice_max(TimeFrame, n = 1)

list(
    # the title will include the year, so if we update the data
    # the dashboard will stay correct
    title = paste("# live births,", most_recent_year$TimeFrame),
    # scales::comma adds commas after the thousands, millions, etc.
    value = scales::comma(most_recent_year$Data)
)
```

### Total children

```{r}
#| content: valuebox
# ^ valuebox will display this as a number in a box

# these data are from the American Community Survey
ages = read_csv("data/ages.csv") |>
    # filter to most recent year
    filter(year == max(year))

children = ages |>
    # filter to most recent year
    filter(age_to < 18) |>
    summarize(count=sum(count), year=first(year))

list(
    # the title will include the year, so if we update the data
    # the dashboard will stay correct
    title = paste("# of children in NC,", children$year),
    # scales::comma adds commas after the thousands, millions, etc.
    value = scales::comma(children$count)
)
```

### Children under 5

```{r}
#| content: valuebox
under5 = ages |>
    filter(age_to < 5) |>
    summarize(count = sum(count), year=first(year))

children = ages |>
    # filter to most recent year
    filter(age_to < 18) |>
    summarize(count=sum(count), year=first(year))

list(
    # the title will include the year, so if we update the data
    # the dashboard will stay correct
    title = paste("% of children under 5,", under5$year),
    # scales::percent formats a proportion as a percentage
    value = scales::percent(under5$count / children$count)
)
```

# Placement instability

<!-- because we did not specify an orientation, this will be row-oriented -->

## Introduction

This page shows how many placements children go through in their first year of care. More placements lead to less stability for children.

## {.tabset}

```{r}
#| title: Trends
#| fig-alt: Bar plot showing the number of placements children go through in their first year of care, showing that 25-20% have only one, and 30-40% have four or more, though this reduced in 2023.

# fig-alt specifies text to be shown or read when the image is not available or accessible. This will most often be used
# by users with visual impairments who use screen reading software. This is a requirement for most public sector websites.

placement_counts = read_csv("data/placement_count_by_year.csv") |>
    filter(Location == "North Carolina") |>
    # this instructs R to display the number of placements data in this order
    mutate(DataGroups = ordered(DataGroups, levels=c("One", "Two", "Three", "Four or More")))

# The NC Child dashboard displays this as four lines on a line plot, but it is hard to see overall trends
# that way. I am going to display this as a stacked bar chart instead.
placement_counts |>
    # ggplot is (imho) the best plotting library out there. gg stands for the "grammar of graphics" -
    # you map aspects of your data onto aspects of your plot; these are known as "aesthetics" (aes for short)
    ggplot(aes(x=TimeFrame, y=Data, fill=DataGroups)) +
        # geoms are what actually display your data - here, a stacked column plot
        # position_fill is a stacked column plot where the bars are scaled to fill the y axis
        # our data roughly total 100%, but aren't perfect due to rounding
        geom_col(position = position_fill(reverse=TRUE)) +
        scale_y_continuous(labels = scales::percent) +
        xlab("Year") +
        ylab("") +
        labs(fill = "Number of Placements") +
        theme(legend.position = "bottom")
```

```{r}
#| title: Table
# The NC Child Portal only displays the most recent year, so we'll do that as well.
# because we converted to an ordered factor above, we'll get the rows in order, unlike
# the NC Child dashboard

placement_counts |>
    filter(TimeFrame == max(TimeFrame)) |>
    # this selects just the two columns we want in our table
    select(DataGroups, Data) |>
    # This adds a percentage sign
    mutate(Data = paste0(Data, "%")) |>
    # This turns the data into an HTML table
    gt() |>
        cols_label(
        DataGroups ~ "Number of Placements",
        Data ~ "Percent of Children"
        )
```

##

_Source: NC Child/NC Department of Health and Human Services (DHHS) in partnership with the UNC Jordan Institute for Families_

# Time in system

The goal of the foster care system is to place children in permanent, safe homes as quickly as possible. This page shows trends in placement speed.

## {.tabset}

```{r}
#| title: Trends
#| fig-alt: Plot of number of children still in the foster care system two years after entering it. It shows a slightly increasing trend, from about 27% in 2011 to about 40% in 2023.


# Setting LocationID = col_character will read the location ID as text, instead of as a number
# which will be important for joining with other data.
two_years = read_csv("data/in_system_two_years.csv", col_types=cols(LocationID=col_character()))

two_years |>
    filter(Location == "North Carolina") |>
    ggplot(aes(x = TimeFrame, y = Data)) +
        geom_line() +
        xlab("Year") +
        # \n will become a line break
        ylab("Percent of children still in\nfoster system after two years") +
        ylim(0, 40)
```

```{r}
#| title: Table

two_years |>
    filter(Location == "North Carolina") |>
    select(TimeFrame, Data) |>
    mutate(Data = paste0(Data, "%")) |>
    gt() |>
        cols_label(
            TimeFrame ~ "Year",
            Data ~ "Percent of children still in foster system after two years"
        )
```

```{r}
#| title: Map
#| fig-width: 8
#| fig-height: 3
#| fig-alt: Map of two-year retention, showing retention in lowest in the central parts of the state and highest in the western part.
#| fig-format: png

# I set custom fig-height and fig-width attributes above to make the map more readable. This is a trial and error
# situation, and counterintuitively making the numbers smaller can often make things more readable, because
# the smaller image is then scaled up. For some reason fig-format: svg does not work with maps, so we change it to png.

# we need to join the county level data for the most recent year with the spatial data so we can map it
two_years_spatial = two_years |>
    filter(TimeFrame == max(TimeFrame)) |>
    # match the percent of children data to the counties data, matching the LocationID column
    # in the children data to the GEOID column in the counties data
    inner_join(counties, by=c("LocationID"="GEOID"))

ggplot() +
    layer_spatial(two_years_spatial, aes(fill=Data), linewidth=0) +
    labs(fill = "Percent of children still in\nfoster system after two years") +
    # this just removes all plot elements except the map
    theme_void()
```

##

_Source: NC Child/NC Department of Health and Human Services (DHHS) in partnership with the UNC Jordan Institute for Families_

<!-- Exercise: add a final tab "Permanency" here showing trends, a table, and a map of permanent placements after one year (data/permanency_rate.csv).
    The data for the other sections on the NC Child Dashboard are also in the data/ folder is you want to do more. -->